---
title: "applied_ps1_new"
author: "Rabail Sofi"
date: "2023-04-05"
output:
  pdf_document: default
  html_document: default
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
rm(list=ls())
library(tinytex)
library(tidyverse)
library(devtools)
library(palmerpenguins)
library(ggthemes)
library(rmarkdown)
library(knitr) 
library(dplyr)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```


# Front matter

This submission is my work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: \<\<\ RS \>\>

Add your collaborators: \<\<\_\_\>\>

Late coins used this pset: 0. Late coins left: 4. <!--You may use one for a given assignment.-->

# R for Data Science Exercises


## First Steps

## Flight Data: Part 1

## Download BTS Data 

1.1.1
```{r}
setwd("/Users/rabailsofi/Desktop/Grad School /Q3, 2023/Data & Programming in R/Week 3/CSVs")

```


1.1.2
```{r}
# repo called "data" saved.
```


1.1.3
```{r}
download.file("https://github.com/datasci-harris/Applied-PS1-Data-Files/blob/main/On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_1.zip", destfile = "rabailsofi/Desktop/data/On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_1.zip")
```


1.1.4
```{r}
system.time(download.file("https://github.com/datasci-harris/Applied-PS1-Data-Files/blob/main/On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_1.zip", destfile = "rabailsofi/Desktop/data/On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_1.zip"))
```

1.1.5 
```{r}

website <- "https://github.com/datasci-harris/Applied-PS1-Data-Files/blob/main/On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_" 
                            
path <- str_c("/Users/rabailsofi/Desktop/data",".zip") 
  
files <- str_c(1:12, ".zip")

download.file(url = str_c(website, files), destfile = str_c(path, files))

```

1.1.6 
```{r}
setwd("~/Desktop")

flights1 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_1.csv")
flights2 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_2.csv") 
flights3 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_3.csv") 
flights4 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_4.csv") 
flights5 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_5.csv") 
flights6 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_6.csv") 
flights7 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_7.csv") 
flights8 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_8.csv") 
flights9 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_9.csv") 
flights10 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_10.csv") 
flights11 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_11.csv") 
flights12 <- read.csv("On_Time_Reporting_Carrier_On_Time_Performance_1987_present_2016_12.csv") 

flights1 <- flights1 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights2 <- flights2 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights3 <- flights3 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights4 <- flights4 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights5 <- flights5 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights6 <- flights6 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights7 <- flights7 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights8 <- flights8 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights9 <- flights9 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights10 <- flights10 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights11 <- flights11 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")
flights12 <- flights12 |>
  filter(OriginState == "IL" | DestStateName == "Illinois")

```

1.1.7
```{r}
allflightsdata <- bind_rows(flights1, flights2, flights3, flights4, flights5, flights6, flights7, flights8, flights9, flights10, flights11, flights12)   

```

1.1.8
```{r}
read_csv("allflightsdata.csv")
```

1.1.9
```{r}
# Manually knitted PDF, timed up to 2 minutes and 10 seconds. 
```

1.1.10
```{r}
view("allflightsdata.csv")
```

1.1.11
```{r}
# *.csv
```

## Data Description 

1.2.1
The tail numbers and the dates. 

1.2.2

1.) The print() function and head() function are almost identical. However, print() gives a better output because you can still scroll through all the columns and rows while previewing the data, while head() doesn't allow us to scroll and view more data. 

2.) The summary() function gives us a better output because its able to give us the averages, minimums, and maximums of each variable in the data set. The glimse() function also gives a good output because it shows us a quick preview of the data and categories in each variable. 
```{r}
print(allflightsdata)
head(allflightsdata)

glimpse(allflightsdata)
str(allflightsdata)
View(allflightsdata)
summary(allflightsdata)
```


## Data Cleaning 

1.3.1
110 variables. 

1.3.2 
```{r}
nycflights <- nycflights13::flights 
view(nycflights)
?flights
```

1.3.3
Variables from nycflights such as air time and time hour were missing in the newly downloaded data set. 

## Data Validation 

1.4.1

The test_that function does not work on my end, but my newly compiled data set, allflightsdata, does have a total of 675822 rows. 
```{r}
library(testthat)
test_that(
  "we have the right number of rows",
  expect_equal(nrow(allflightsdata), 675822))
```

1.4.2
```{r}

test_that(
  "we have the right number of rows",
  expect_equal(nrow(allflightsdata$OriginState=="IL"), 340625))

allflightsdata |>
  filter(OriginState=="IL")

```

1.4.3

After removing flights to and from ORD and MDW, there are only 12,240 flights left in the dataframe. 

Citation: https://stackoverflow.com/questions/6650510/remove-rows-from-data-frame-where-a-row-matches-a-string 
```{r}
exclude_midway_ohare <-allflightsdata[(!(allflightsdata$Origin=="ORD") & !(allflightsdata$Origin=="MDW") & !(allflightsdata$Dest=='ORD') & !(allflightsdata$Dest=="MDW")),]
```

1.4.4
```{r}

top_5_origins <- exclude_midway_ohare |>
  count(Origin) |>
  slice_max(n, n=5) |>
  print(top_5_origins)

top_5_dests <- exclude_midway_ohare |>
  count(Dest) |>
  slice_max(n, n=5) |>
  print(top_5_dests)

```

1.4.5

1.) When we limit the flights only to and from ORD and MDW, the number of flights should be 663582 (total number of flights, 675822 - flights without ORD and MDW, 12240.
```{r}
include_midway_ohare <- allflightsdata |>
  filter(Origin=="ORD" | Origin=="MDW" | Dest=="ORD" | Dest=="MDW") 
```

1.4.6
According to an article by Upgraded Points, Ohare and Midway are "a major airport for connections all over the U.S., operating as the main hub for United Airlines and the third-largest hub for American Airlines". Luftansa and Air Canada are next largest airlines that enter and exit Ohare and Midway. While real time data of how many flights enter and exit those two airports isn't avaible, sources say that north of 900 flights enter & exit the city every single day. Knowing that Chicgao is one of the most visited cities in the United States, these numbers and the BTS data seem to match and this dataset can be assessed as believable. 

Citation: https://upgradedpoints.com/travel/airports/chicago-ohare-ord-airport/ & https://simpleflying.com/chicago-o-hare-world-top-regional-jet-airport/#:~:text=Up%20To%20914%20Flights%20A,World's%20Top%20Regional%20Jet%20Airport

## Flight Data: Part II

2.1.1

November has the lowest average arrival delays. At least 80% of the flights are on time in October as per the table 'flights_on_time_Chicago'. In the ggplot below, we can see that October and November have the lowest average arrival delays while also most of the flights during these months are on time. Some other months that have flights on time are January and February. In terms of average arrival delays, the summer months have the most delays and the colder months have fewer delays. 

Citation: https://www.statology.org/ggplot-multiple-data-frames/ & https://r-graph-gallery.com/line-chart-dual-Y-axis-ggplot2.html 
```{r}
summary_table_Chicago <- allflightsdata |>
  filter(DestCityName=="Chicago, IL") |>
  group_by(Month) |> 
  summarize(avg_arr_delay = mean(ArrDelay, na.rm = TRUE),
            Flights_On_Time = mean(ArrDelay <=0, na.rm= TRUE) * 100,
            n=n()) 

include_midway_ohare_1<- include_midway_ohare
include_midway_ohare_1$ArrDelayMinutes <- ifelse(include_midway_ohare$ArrDelay > 0, '1', '0')

flights_on_time_Chicago<- include_midway_ohare_1 |>
  group_by(Month) |>
  summarize(ave_arr_delay = mean(ArrDelay, na.rm = TRUE))
view(flights_on_time_Chicago)

ggplot(summary_table_Chicago, aes(x=Month, y=avg_arr_delay, size = Flights_On_Time)) +
   geom_line(color="#69b3a2", size=2) +
   ggtitle("Average Flight Delays in Chicago") +
   geom_point() 

```

2.1.2

As seen in the ggplot below, the summer months (July to August) are when we see the most flights going to Chicago. The number of flights peak in July with a count of 60700 and August has the next highest number of flights with a count of 60136. We see the lowest dips during February and January. 
```{r}

flights_count_Chicago <- include_midway_ohare_1 |>
  group_by(Month) %>%
  count(Flights)
 avg_dep_delay_Chicago <- include_midway_ohare_1 |>
  group_by(Month) |>
  summarize(ave_dep_delay = mean(DepDelay, na.rm=TRUE))


ggplot(flights_count_Chicago, aes(x=Month, y=n)) +
  geom_line(color="#69b3a2", size=2) +
  ggtitle("Flights to Chicago by Month") +
  geom_point() 
```

2.1.3

The convention should be held between October and November because those month have some of the largest counts of flights going to Chicago, but don't have as long delays (arrivals as well as departures) as more common months such as July or August. 

a.) 
```{r}
include_midway_ohare_1 <- include_midway_ohare
colnames(include_midway_ohare_1)[44] ="FlightsOnTime"

summary_table_include_midway_ohare_1 <- include_midway_ohare_1 |>
  group_by(Month) |>
  summarize(Ave_Arrival_Delays = mean(ArrDelay, na.rm = TRUE),
            Ave_Departure_Delays = mean(DepDelay, na.rm = TRUE),
            Flights_On_Time = mean(ArrDelay <=0, na.rm= TRUE) * 100,
            n=n()) 

ggplot_1<- ggplot(summary_table_include_midway_ohare_1, aes(Month, Flights_On_Time, size = n))+
  labs(subtitle="Number of Flights & Flights on Time", 
       y="Flights On Time", 
       x="Months", 
       title="Top Months For Chicago's 2-Day Convention in 2024", 
       caption = "Source: Source: Bureau of Transportation Statistics")+
  geom_point()+
  labs(color = "Flight Count") +
  guides(size = guide_legend(title = "Flight Count"))
print(ggplot_1)

```


b.) In the richer plot, I added the average departure delays to the plot to further prove that October and November indeed have some of the lowest departure delays and would ideal options for the convention in 2024. Adding the geom_smooth allows us to visualize the outliers and in this case, its evident that October & November have extraordinary punctual flights and the flights during this time is in high volume still. 

Citation: https://www.geeksforgeeks.org/how-to-create-tables-in-r/
```{r}
ggplot_2 <- summary_table_include_midway_ohare_1 |>
  ggplot(aes(Month, Flights_On_Time, size =n, color = Ave_Departure_Delays)) +
  labs(y="Flights On Time", 
       x="Months", 
       title="Top Months For Chicago's 2-Day Convention in 2024", 
       caption = "Source: Bureau of Transportation Statistics")+
    geom_point() + geom_smooth(method = "lm")+
  guides(size = guide_legend(title = "Flight Count"))
print(ggplot_2)

```

c.) I would submit the plot titled 'ggplot_2' because its more descriptive and visualizes the distribution within the data better. 

d.) The other data sets that I would use "tourism.rds" and "OpenFlights" which tells us about flights and tourism. These data sets help us with planning the date for the convention. 

## Pick An Airline 

2.2.1
1.) Now that the recommended month is October, we can plot the most frequent airlines that fly to and from Chicago. Based on the plot created below, ggplot3, we can suggest the top airlines used in October which is United (UA) and SouthWest (WN). 

```{r}
ggplot3 <- include_midway_ohare_1 |>
   ggplot(mapping = aes(x = Month, fill = Reporting_Airline)) +
   geom_bar(position = "dodge")+
   labs(y= "Flight Count",
        x="Month",
        title="Airlines For Chicago's Convention 2024", 
        caption = "Source: Bureau of Transportation Statistics")+
   guides(size = guide_legend(title = "Airlines"))
print(ggplot3)
```
## Reconsider the date

2.3.1
1.) Now that the Convention has been moved to Greer, SC, we can see that the best time to host the event is February. In the plot, ggplot4, and in the summary table titles 'summary_table_include_south_carolina',  we can see that February has the highest number of flights on time, has some of the largest volumes of flights, and has relative fewer delays than some of it busier months. 

```{r}
include_south_carolina <- allflightsdata |>
  filter(OriginCityName=="Greer, SC" | DestCityName=="Greer, SC") 


summary_table_include_south_carolina <- include_south_carolina |>
  group_by(Month) |>
  summarize(Ave_Arrival_Delays = mean(ArrDelay, na.rm = TRUE),
            Ave_Departure_Delays = mean(DepDelay, na.rm = TRUE),
            Flights_On_Time = mean(ArrDelay <=0, na.rm= TRUE) * 100,
            n=n()) 
ggplot_4 <- summary_table_include_south_carolina |>
  ggplot(aes(Month, Flights_On_Time, size =n, color = Ave_Departure_Delays)) +
  labs(y="Flights On Time", 
       x="Months", 
       title="Top Months For Greer's 2-Day Convention in 2024", 
       caption = "Source: Bureau of Transportation Statistics")+
    geom_point() + geom_smooth(method = "lm")+
  guides(size = guide_legend(title = "Flight Count"))
print(ggplot_4)

```






